Forecasting the US Presidential Election: A Poll-of-Polls Approach Using Linear Models

### 2024-10-22 paper/paper.qmd
save spline fit as models, update paper in model section

@@

Our data is download from [Five Thirty Eight](https://projects.fivethirtyeight.com/polls)[@polls], the website gathered survey data from different pollsters of 2024 US president election, We use the statistical programming **language R** [@citeR], and packages **lubridate**[@lubridate], **dplyr**[@dplyr], **tidyverse**[@tidyverse], **model summary**[@models] to process the data. Also, we use **tibble** [@tibble] to create the table for simulate table and save it as csv files. Graphs are created by using ggplot2 in from package **tidyverse**[@tidyverse]Following **Telling stories with data**[@telling_stories_with_data], we consider using multiple linear model to predict and forecasting the result of US president election in 2024. We are creating different graphs and models for the candidates from different parties in general state and combine all models and graphs together to create a new scatter plot to predict and forecasting the US president election in 2024.

@@

## Outcome variables

@@

## Predictor variables

@@

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Our model is build from a transformed version of the MLR in above section, with transparency_score dropped as it has a significant level greater than 0.05.

@@

\begin{align} 
y_i | p_i &\sim \mbox{Binomial}(\text{sample\_size}_i, p_i) \\
\log \left( \frac{p_i}{1 - p_i} \right) &= \beta_0 + \beta_1 \log(\text{sample\_size}_i) + \beta_2 \, \text{pollscore}_i + \beta_3 \, \text{numeric\_grade}_i + \alpha_j \\
\alpha_j &\sim \mbox{Normal}(0, \sigma_{\text{pollster}}) \\
\beta_0 &\sim \mbox{Normal}(0, 5) \\
\beta_1 &\sim \mbox{Normal}(0, 5) \\
\beta_2 &\sim \mbox{Normal}(0, 5) \\
\beta_3 &\sim \mbox{Normal}(0, 5) \\
\sigma_{\text{pollster}} &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We first priors as default with “autoscale = TRUE”,

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

library(rstanarm)

# first_model <-
#   readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

# modelsummary::modelsummary(
#   list(
#     "First model" = first_model
#   ),
#   statistic = "mad",
#   fmt = 2
# )
```

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

@@

# Model details {#sec-model-details}

>>>>

cleaned_data <- read_csv(here::here("data/02-analysis_data/cleaned_data.csv"))
nationwide_data <- cleaned_data %>%  filter(state == "National")

# only two entries for Conservative Party
unique_parties <- unique(nationwide_data$party)

# drop columns that have NA in numeric_grade or pollscore, and 2 entries for CON party
nationwide_data <- nationwide_data %>%
  drop_na(numeric_grade, pollscore) %>%
  filter(party != "CON") %>%
  mutate(days_since_start = as.numeric(end_date - min(end_date)))

unique_parties <- unique(nationwide_data$party)
nationwide_data$party <- as.factor(nationwide_data$party)

nationwide_data <- nationwide_data %>% mutate(
  num_party = round((pct / 100) * sample_size, 0)
)
# filter nationwide_data by parties
dem_data <- nationwide_data %>%
  filter(party == "DEM")
dem_data <- na.omit(dem_data)

dem_data <- dem_data |>
  mutate(
    pollster = factor(pollster)
  )

rep_data <- nationwide_data %>%
  filter(party == "REP")
rep_data <- na.omit(rep_data)

rep_data <- rep_data |>
  mutate(
    pollster = factor(pollster)
  )
```

@@

Our data is download from [@polls], the website gathered survey data from different pollsters, We use the statistical programming language R [@citeR], and packages [@lubridate], [@dplyr], [@tidyverse], to process the data. Following @tellingstories, we consider...

@@

## Outcome variables {#sec-octvar}

@@

## Predictor variables {#sec-prevar}

@@

# Models

The goal of our modelling strategy is try to capture the. Firstly,...
In the following section we briefly describe the models we used to investigate. Background details and diagnostics are included in [Appendix -@sec-model-details].

## SLR Model set-up

## MLR Model set-up
We first filtered our data by different parties, as the single linear model but include other 3 parties(GRE, IND, and LIB, Conservative only has two entries in cleaned nationwide data and has 0 pct so its been removed), and fit linear model for each of them. The formula for Multiple linear regression is as follow
\begin{align}
\text{pct}_i &= \beta_0 + \beta_1 \log(\text{sample\_size}_i) + \beta_2 \, \text{pollscore}_i + \beta_3 \, \text{numeric\_grade}_i \\
&+ \beta_4 \, \text{transparency\_score}_i + \beta_5 \, \text{days\_since\_start}_i + \epsilon_i
\end{align}

Response Variable is pct. Predictors are log(sample_size), pollscore, numeric_grade, transparency_score, and days_since_start, detail can be found in [@sec-octvar], and [@sec-prevar]

### Result

Our results for the MLR model are summarized in @tbl-mlrmodelresults.

MLR_models <-
  readRDS(file = here::here("models/MLRmodels.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-mlrmodelresults
#| tbl-cap: "Explanatory models of pct based on pollscore, sample size, and time"
#| warning: false

modelsummary::modelsummary(
  MLR_models,
  fmt = 2
)
```
We first check the residual vs. fitted plot as shown in [Appendix @sec-model-details], and see no obvious violation of assumptions. According to the summary, DEM and REP models are based on the largest number of observations (both 1235). Smaller parties have fewer observations, which may lead to less robust models. The $r^2$ shows that the model explain the highest amount of variance on Democratic Party than others with value 0.357, and other parties less than 0.2. Based on the summary of models for DEM and REP [@tbl-mlrmodelsum], we are not sure if those predictor that measure quality of polls are significant or not, so we keep them for now in Bayesian model. Prediction of MLR models is shown in [@fig-mlrpredict], which we can see blue line which is Democratic party has a steeper slope and start to take the lead between day 100 and day 200.

```{r}
#| echo: false
#| label: fig-mlrpredict
#| fig-cap: "Prediction with MLR models for DEM and REP party"
#| warning: false

# Augment the datasets with predictions
dem_data <- dem_data |>
  mutate(fitted_pct_dem = predict(MLR_models[['DEM']]))
rep_data <- rep_data |>
  mutate(fitted_pct_rep = predict(MLR_models[['REP']]))

# Combine both datasets into one
combined_data <- bind_rows(dem_data, rep_data)

# Plot with colors based on the party
ggplot(combined_data, aes(x = days_since_start)) +
  geom_line(aes(y = fitted_pct_dem, color = "DEM")) +
  geom_line(aes(y = fitted_pct_rep, color = "REP")) +
  scale_color_manual(values = c("DEM" = "blue", "REP" = "red"), 
                     name = "Party",
                     labels = c("DEM (Blue)", "REP (Red)")) +
  theme_classic() +
  labs(y = "Percent", x = "Days") +
  theme(legend.position = "top")
```

## Bayesian Model set-up
The Bayesian model is build from the MLR models from above section. With the inspiration from the example R code, we introduce the random effects for Pollsters, and we use a logistic link to predict the probability of support, modeling count data which is in binomial distribution. We choose to use a default prior and enable autoscale, the formula is shown below

\begin{align} 
\log \left( \frac{p_i}{1 - p_i} \right) &= \beta_0 + \beta_1 \log(\text{sample\_size}_i) + \beta_2 \, \text{pollscore}_i + \beta_3 \, \text{numeric\_grade}_i \\
&+ \beta_4 \, \text{transparency\_score}_i + \beta_5 \, \text{days\_since\_start}_i + \alpha_j
\end{align}
\alpha_j &\sim \text{Normal}(0, \sigma_{\text{pollster}}) \\
\beta_0 &\sim \text{Normal}(0, 2.5) \\
\beta_1 &\sim \text{Normal}(0, 2.5) \\
\beta_2 &\sim \text{Normal}(0, 2.5) \\
\beta_3 &\sim \text{Normal}(0, 2.5) \\
\beta_4 &\sim \text{Normal}(0, 2.5) \\
\beta_5 &\sim \text{Normal}(0, 2.5) \\
\sigma_{\text{pollster}} &\sim \text{Exponential}(1)
\end{align}

@@

### Result
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

Bay_model <-
  readRDS(file = here::here("models/bayesian_model_dem.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-baymodelresults
#| tbl-cap: ""
#| warning: false

modelsummary::modelsummary(
  Bay_model,
  fmt = 2
)
```

# Prediction

To get a prediction by the Vayesian model we have, we first spline fit the model, and then create a data frame and generate posterior predictions by the new data frame.
## Spline Fit for Bayesian model set-up

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use priors $\text{Normal}(0, 5) \\$ to allow for more flexibility of predictors effects with “autoscale = TRUE”. Then we crate a data frame like [@tbl-newdata], we set the quality scale of predict data to the best to perform a poll that has high quality. the result is shown as [@sec-splresult]
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-newdata

# Create new data for prediction
new_data <- data.frame(
  pollscore = -1.5,
  sample_size = 1200,
  numeric_grade = 3,
  transparency_score=10,
  days_since_start = seq(
    min(dem_data$days_since_start),
    max(dem_data$days_since_start),
    length.out = 100
  ),
  pollster = factor("TIPP", levels = levels(dem_data$pollster))
)

head(new_data)
```

## Result {#sec-splresult}
```{r}
#| echo: false
#| warning: false
#| message: false

spline_dem <-
  readRDS(file = here::here("models/spline_dem.rds"))
spline_rep <-
  readRDS(file = here::here("models/spline_rep.rds"))
```

```{r}
#| echo: false
#| label: fig-splpredict
#| fig-cap: "Poll Percentage over Time with Spline Fit for DEM and REP party"
#| warning: false

posterior_preds_dem <- posterior_predict(spline_model_dem, newdata = new_data)
posterior_preds_rep <- posterior_predict(spline_model_rep, newdata = new_data)

combined_data <- bind_rows(dem_data, rep_data)

# Plot the spline fit
ggplot(combined_data, aes(x = days_since_start, y = pct, color = pollster)) +
  # DEM predictions
  geom_line(
    data = pred_summary %>% filter(party == "DEM"),
    aes(x = days_since_start, y = pred_mean_dem, color = "DEM"),  # Move color inside aes()
    inherit.aes = FALSE
  ) +
  geom_ribbon(
    data = pred_summary %>% filter(party == "DEM"),
    aes(x = days_since_start, ymin = pred_lower_dem, ymax = pred_upper_dem, fill = "DEM"),  # Move fill inside aes()
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  # REP predictions
  geom_line(
    data = pred_summary %>% filter(party == "REP"),
    aes(x = days_since_start, y = pred_mean_rep, color = "REP"),  # Move color inside aes()
    inherit.aes = FALSE
  ) +
  geom_ribbon(
    data = pred_summary %>% filter(party == "REP"),
    aes(x = days_since_start, ymin = pred_lower_rep, ymax = pred_upper_rep, fill = "REP"),  # Move fill inside aes()
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  # Use scale_color_manual and scale_fill_manual for colors and fills
  scale_color_manual(values = c("DEM" = "blue", "REP" = "red"), 
                     name = "Party",
                     labels = c("DEM (Blue)", "REP (Red)")) +
  scale_fill_manual(values = c("DEM" = "blue", "REP" = "red"), 
                    name = "Party",
                    labels = c("DEM (Blue)", "REP (Red)")) +
  labs(
    x = "Days since 2024",
    y = "Percentage",
  ) +
  theme_minimal()
```
The prediction graph contains prediction of spline fit model for two parties, blue line represents the DEM party's predicted percentage of support, and the red line represents the REP party's predicted percentage of support over time (measured in days since the start of 2024). We can tell that the predict support of DEM party is taking the lead by aroud 2 percent until the last day of the dataset.

@@

## Crossing point in the prediction  {#sec-first-point}

From the prediction of spline fit model [@fig-splpredict], There is a clear increase in the support for the DEM party starting around day 200, while the REP party's support remains more stable with only a slight increase over time.
The increase in support for the DEM party in the predictions could be attributed to Harris's rise as the Democratic candidate after Biden's exit from the race On July 21, 2024, which is about 230 days from start of 2024.

@@

# Model details {#sec-model-details}
## Assumption check for MLR models
```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-assumptioncheck
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: [""]

for (party in names(MLR_models)) {
  # Extract the model for the current party
  current_model <- MLR_models[[party]]

  # Create a Residuals vs Fitted plot for the current model
  plot(current_model, which = 1, main = paste("Residuals vs Fitted for", party, "Party"))

}
```

## Significant check for MLR models
```{r}
#| echo: false
#| eval: true
#| label: tbl-mlrmodelsum
#| tbl-cap: "Summary of DEM and REP data fit by MLR model"
#| warning: false

summary(MLR_models[['DEM']])
summary(MLR_models[['REP']])
```
